### 2013-06-30 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
adding full clip scaling section
adding asynchronous algorithms for neumann ulam transport

In this section Brunner and Brantley's fully asynchronous algorithm is
briefly presented along with its application to the Neumann-Ulam
method. Their algorithm is presented in detail in the reference and
was effectively implemented verbatim for this work. In their work they
identify two data sets that are required to be communicated: the
sharing of particles that are transported from one domain to another
and therefore from one processor to another and a global communication
that signals if particle transport has been completed on all
processors. Both of these processes will be addressed at a high level
along with how the parallel Monte Carlo data structures are
constructed.

@@

\subsection{Domain-to-Domain Communication Kernel }
\label{subsec:domain_to_domain_kernel}

@@

\subsection{Ending the Transport Sequence }
\label{subsec:ending_transport}

@@

%%---------------------------------------------------------------------------%%
\section{Multiple-Set Overlapping-Domain Algorithm\ }
\label{subsec:msod}
Although the implementation presented in the previous section was
observed by Brunner and Brantley to be robust and allowed for scaling
to large numbers of processors, performance issues were still noted
with parallel efficiency improvements needed in both the weak and
strong scaling cases for unbalanced problems. These results led them
to conclude that a combination of domain decomposition and domain
replication could be used to solve some of these issues. In 2010,
Wagner and colleagues developed the \textit{multiple-set
  overlapping-domain} (MSOD) decomposition for parallel Monte Carlo
applications for full-core light water reactor analysis
\citep{wagner_hybrid_2010}. In their work, an extension of Brunner's,
their scheme employed similar parallel algorithms for particle
transport but a certain amount of overlap between adjacent domains was
used to decrease the number of particles leaving the local domain. In
addition, Wagner utilized a level of replication of the domain such
that the domain was only decomposed on $O(100)$ processors and if
replicated $O(1,000)$ times potentially achieves simulation on
$O(100,000)$ processors, thus providing both spatial and particle
parallelism. Each collection of processors that constitutes a
representation of the entire domain is referred to as a set, and
within a set overlap occurs among its sub-domains. The original
motivation was to decompose the domain in a way that it remained in a
physical cabinet in a large distributed machine, thus reducing latency
costs during communication. A multiple set scheme is also motivated by
the fact that communication during particle transport only occurs
within a set, limiting communications during the transport procedure
to a group of $O(100)$ processors, a number that was shown to have
excellent parallel efficiencies in Brunner's work and therefore will
scale well in this algorithm. The overlapping domains within each set
also demonstrated reduced communication costs. On each processor, the
source is sampled in the local domain that would exist if no overlap
was used while tallies can be made over the entire overlapping domain.

@@

%%---------------------------------------------------------------------------%%
\section{Parallel MCSA\ }
\label{sec:parallel_mcsa}
With the parallel adjoint Neumann-Ulam solver implementation described
aboive, the parallel implementation of the MCSA method is
trivial. Recall the MCSA iteration procedure outlined in
Eq~(\ref{eq:mcsa}). In \S\ref{sec:parallel_krylov_methods} we
discussed parallel matrix and vector operations as utilized in
conventional Krylov methods. We utilize these here for the parallel
MCSA implementation. In the first step, a parallel matrix-vector
multiply is used to apply the split operator to the previous iterate's
solution. A parallel vector update is then performed with the source
vector to arrive at the initial iteration guess. In the next step, the
residual is computed by the same operations where now the operator is
applied to the solution guess with a parallel matrix-vector multiply
and then a parallel vector update with the source vector is
performed. Once the correction is computed with a parallel adjoint
Neumann-Ulam solve, this correction is applied to the guess with a
parallel vector update to get the new iteration
solution. Additionally, as given by
Eq~(\ref{eq:mcsa_stopping_criteria}), 2 parallel vector reductions
will be required to check the stopping criteria: one initially to
compute the infinity norm of the source vector, and another at every
iteration to compute the infinity norm of the residual vector. For
this implementation, all of the issues that will be potentially
generated by the parallel adjoint solver implementation will manifest
themselves here as the quality of the correction will be of intense
study.

@@

@@

>>>>

In this section Brunner and Brantley's fully asynchronous algorithm is
presented along with its application to the Neumann-Ulam method. Their
algorithm is presented in further detail in the reference and was
effectively implemented verbatim for this work. In their work they
identify two data sets that are required to be communicated: the
sharing of particles that are transported from one domain to another
and therefore from one processor to another and a global communication
that signals if particle transport has been completed on all
processors. Both of these processes will be addressed at a high level
along with how the Monte Carlo data structures are
constructed in parallel.

@@

\subsection{The Algorithm}
\label{subsec:parallel_mc_algorithm}

\begin{algorithm}[h!]
  \caption{Parallel Neumann-Ulam Algorithm}
  \label{alg:parallel_mc_algorithm}
  \begin{algorithmic}[1]
    \State get list of neighbor processors 
    \Comment{Each neighbor owns an adjacent subdomain}
    \ForAll{neighbors}
    \State post nonblocking receive for maximum history buffer size
    \State allocate history buffer
    \EndFor
    \State \textit{historiesCompleted} = 0
    \Comment{local+child finished histories}
    \State \textit{localProcessed} = 0
    \Comment{local trajectories computed (not necessarily finished)}
    \State calculate parent and children processor ID numbers in
    binary tree
    \ForAll{child processes}
    \State post nonblocking receive for \textit{historiesCompleted} history tally
    \EndFor
    \State post nonblocking receive for stop message from parent
    \While{stop flag not set}
    \If{any local histories in source or stack}
    \State \textbf{LocalHistoryTransport()}
    \State ++\textit{localProcessed}
    \EndIf
    \If{checkPeriod == \textit{localProcessed} || no local histories}
    \State \textbf{ProcessMessages()}
    \State \textit{localProcessed} = 0
    \EndIf
    \If{no local histories}
    \State \textbf{ControlTermination()}
    \EndIf
    \EndWhile
    \State cancel outstanding nonblocking requests
    \State free all history buffers
  \end{algorithmic}
\end{algorithm}

\subsubsection{Domain-to-Domain Communication Kernels}
\label{subsubsec:domain_to_domain_kernel}

\begin{algorithm}[h!]
  \caption{\textbf{LocalHistoryTransport()}}
  \label{alg:local_history_transport}
  \begin{algorithmic}[1]
    \State transport history through the domain until termination
    \If{history is in neighbor boundary state}
    \State add history to neighbor buffer
    \If{neighbor buffer is full}
    \State nonblocking send history buffer to neighbor
    \State allocate new history buffer for neighbor
    \EndIf
    \Else
    \If{history terminated by weight cutoff}
    \State post-process history
    \State ++\textit{historiesCompleted}
    \EndIf
    \EndIf
  \end{algorithmic}
\end{algorithm}

\begin{algorithm}[h!]
  \caption{\textbf{ProcessMessages()}}
  \label{alg:process_messages}
  \begin{algorithmic}[1]
    \ForAll{received history buffers from neighbor}
    \State unpack number of histories in buffer
    \State add the histories to the running stack
    \State repost nonblocking receive with neighbor
    \EndFor
    \ForAll{\textit{historiesCompleted} messages from children}
    \State \textit{historiesCompleted} += message value
    \State repost nonblocking receive with child
    \EndFor
  \end{algorithmic}
\end{algorithm}

@@

\subsubsection{Ending the Transport Sequence}
\label{subsubsec:ending_transport}

\begin{algorithm}[h!]
  \caption{\textbf{ControlTermination()}}
  \label{alg:control_termination}
  \begin{algorithmic}[1]
    \State nonblocking send an partially full history buffers
    \ForAll{\textit{historiesCompleted} messages from children}
    \State \textit{historiesCompleted} += message \textit{historiesCompleted}
    \State repost nonblocking receive with child
    \EndFor
    \If{MASTER processor}
    \If{\textit{historiesCompleted} == global \# of source histories}
    \State set stop flag
    \ForAll{children}
    \State nonblocking send stop message to child
    \EndFor
    \EndIf
    \Else
    \State nonblocking send \textit{historiesCompleted} to parent
    \State \textit{historiesCompleted} = 0
    \State check for stop signal from parent
    \If{stop signal from parent}
    \ForAll{children}
    \State nonblocking send stop message to child
    \EndFor
    \EndIf
    \EndIf
  \end{algorithmic}
\end{algorithm}

@@

%%---------------------------------------------------------------------------%%
\section{Multiple-Set Overlapping-Domain Algorithm\ }
\label{subsec:msod}
Although the implementation presented in the previous section was
observed by Brunner and Brantley to be robust and allowed for scaling
to large numbers of processors, performance issues were still noted
with parallel efficiency improvements needed in both the weak and
strong scaling cases for unbalanced problems. These results led them
to conclude that a combination of domain decomposition and domain
replication could be used to solve some of these issues. In 2010,
Wagner and colleagues developed the \textit{multiple-set
  overlapping-domain} (MSOD) decomposition for parallel Monte Carlo
applications for full-core light water reactor analysis
\citep{wagner_hybrid_2010}. In their work, an extension of Brunner's,
their scheme employed similar parallel algorithms for particle
transport but a certain amount of overlap between adjacent domains was
used to decrease the number of particles leaving the local domain. In
addition, Wagner utilized a level of replication of the domain such
that the domain was only decomposed on $O(100)$ processors and if
replicated $O(1,000)$ times potentially achieves simulation on
$O(100,000)$ processors, thus providing both spatial and particle
parallelism.

Each collection of processors that constitutes a representation of the
entire domain is referred to as a set, and within a set overlap occurs
among its sub-domains. The original motivation was to decompose the
domain in a way that it remained in a physical cabinet in a large
distributed machine, thus reducing latency costs during
communication. A multiple set scheme is also motivated by the fact
that communication during particle transport only occurs within a set,
limiting communications during the transport procedure to a group of
$O(100)$ processors, a number that was shown to have excellent
parallel efficiencies in Brunner's work and therefore will scale well
in this algorithm. The overlapping domains within each set also
demonstrated reduced communication costs. On each processor, the
source is sampled in the local domain that would exist if no overlap
was used while tallies can be made over the entire overlapping domain.

@@

%%---------------------------------------------------------------------------%%
\section{Parallel MCSA\ }
\label{sec:parallel_mcsa}
With the parallel adjoint Neumann-Ulam solver implementation described
aboive, the parallel implementation of the MCSA method is
trivial. Recall the MCSA iteration procedure outlined in
Eq~(\ref{eq:mcsa}). In \S\ref{sec:parallel_krylov_methods} we discuss
parallel matrix and vector operations as utilized in conventional
Krylov methods. We utilize these here for the parallel MCSA
implementation. In the first step, a parallel matrix-vector multiply
is used to apply the split operator to the previous iterate's
solution. A parallel vector update is then performed with the source
vector to arrive at the initial iteration guess. In the next step, the
residual is computed by the same operations where now the operator is
applied to the solution guess with a parallel matrix-vector multiply
and then a parallel vector update with the source vector is
performed. Once the correction is computed with a parallel adjoint
Neumann-Ulam solve, this correction is applied to the guess with a
parallel vector update to get the new iteration
solution. Additionally, as given by
Eq~(\ref{eq:mcsa_stopping_criteria}), 2 parallel vector reductions
will be required to check the stopping criteria: one initially to
compute the infinity norm of the source vector, and another at every
iteration to compute the infinity norm of the residual vector. For
this implementation, all of the issues that will be potentially
generated by the parallel adjoint solver implementation will manifest
themselves here as the quality of the correction will be of intense
study.

@@

How effectively the presented algorithms parallelize MCSA will be
determined by several performance measures. For this work, these
measures come directly from Keyes' primer on the scalability of domain
decomposed methods and specifically his discussion on how to measure
their parallel performance \citep{keyes_how_1999}. All of these
metrics are derived from two measured quantities: the number of
iterations required for convergence and the total wall time required
for convergence. Each of these quantities are then measured in varying
parameter studies by changing the number of processes in a simulation
and by varying the local or global problem size. Using the notation of
Keyes, we will define \textit{parallel efficiency} as $\eta(N,P)$
where the efficiency is a function of the global problem size, $N$,
and the number of processes in the calculation, $P$. There are several
types of efficiencies that will be useful in our analysis of
scalability with different definitions depending on the type of
parametric scaling study performed. For all efficiencies, a value of 1
is considered perfect while a value of 0 signals the worst possible
performance. Often, as was observed in this work, it is the case that
efficiencies above 1 are computed in certain scaling studies. For the
data presented in the following sections, an explanation of each
observed case of superunitary efficiencies will be provided.

Of primary interest to the nuclear engineer is the \textit{strong
  scaling} performance of an algorithm
\citep{siegel_analysis_2012}. In this type of scaling study, the
global problem size is fixed and the local problem size changes with
the number of processes in the calculation. In this case, as $P$
increases, $N$ decreases, thus reducing performance by having a larger
ratio of communication to computational work. This type of scaling is
important because typically the global problem size is fixed due to
the problem of interest. Consider the desire for a high fidelity
neutronics simulation of the entire core of a nuclear reactor. In this
case, the global problem is known from the geometry definition
including the number of fuel assemblies and the geometry of those
assemblies all the way down to the resolution of the space, angle, and
energy discretizations required in order to capture physical phenomena
of interest. Because the global problem size is known a priori, strong
scaling is an important measure of how useful a solution technique is
when larger and larger values of $P$ are used to solve the
problem. Ideally, the time to solution for the the problem will
decrease linearly as a function of $P$, permitting a faster time to
solution using the maximum available resources thus making sure all
$P$ are positively affecting the runtime of the problem.

For strong scaling, the objective is to then decrease the runtime of a
fixed size global problem as a linear function of $P$. We can then
define the strong scaling \textit{absolute efficiency} as:
\begin{equation}
  \eta_{strong}(N,P) = \frac{1}{P} \frac{T(N,1)}{T(N,P)}\:,
  \label{eq:strong_scaling_absolute}
\end{equation}
where $T(N,P)$ is the wall time for a computation of global size $N$
using $P$ processes. Note here that this particular definition is
measured with respect to a serial computation. On leadership class
machines, it is often the case that a problem size large enough to
effectively execute on $O(10,000)$ cores or larger is significantly
larger than any problem that may be solved by a single core due to
memory restrictions. Therefore, we consider absolute scaling for a
base case where $Q$ processes are used instead of a serial
computation:
\begin{equation}
  \eta_{strong}(N,P|Q) = \frac{Q}{P} \frac{T(N,Q)}{T(N,P)}\:.
  \label{eq:strong_scaling_absolute_ref}
\end{equation}
The absolute efficiency measure presented here only considers the
total time to solution. Often, that total time to solution may be
changing for an iterative method due to the fact that the number of
iterations required to converge may be changing as a function of
$P$. This will often happen in cases where modifying $N$ changes the
spectral character of the problem, often making it stiffer and
therefore more difficult to solve when $N$ increases. Therefore we
also consider \textit{algorithmic efficiency} from Keyes' work which
considers this potential change in iteration count:
\begin{equation}
  \eta_{alg}(N,P|Q) = \frac{I(N,Q)}{I(N,P)}\:,
  \label{eq:algorithmic_efficiency}
\end{equation}
where $I(N,P)$ is the number of iterations required to converge a
problem of global size $N$ computed using $P$ processes. We can then
adjust the absolute efficiency using this new algorithmic efficiency
in order to compute the parallel scalability of a single iteration
such that algorithmic effects may be neglected. We define the
\textit{implementation efficiency} as:
\begin{equation}
  \eta_{impl}(N,P|Q) = \frac{\eta}{\eta_{alg}}\:,
  \label{eq:implementation efficiency}
\end{equation}
using the absolute efficiency in the numerator, giving an effective
measure of scalability of a single iteration of the algorithm.

In addition to strong scaling, in certain cases it is useful to
consider the \textit{weak scaling} of an iterative method. In this
case, the ratio $N/P$ is fixed such that the local problem size
remains the same for all values of $P$. In this case, the objective is
to maintain a constant wall time while both $N$ and $P$ grow as a
fixed ratio. In this case, we define the absolute efficiency as:
\begin{equation}
  \eta_{weak}(M|N,P|Q) = \frac{T(M,Q)}{T(N,P)}\:,
  \label{eq:weak_scaling_absolute}
\end{equation}
subject to the constraint $M/Q = N/P$. The scaling is perfect if the
runtime is static for all $P$ when $N/P$ is fixed. Often, weak scaling
is of interest to the nuclear engineer due to the fact that the
computational resources available are fixed and may not be able to
contain the entire global problem of interest. In this case, an
algorithm with good weak scaling performance permits all of the
computational resources available to be used effectively to allow for
the highest fidelity problem possible to be solved. In addition, for
the resources available, the memory to processor ratio on a
distributed machine is often fixed, giving an exercise in fixed $N/P$
to solve the highest fidelity problem possible. Therefore, good weak
scaling should not be viewed as an enhancement of time to solution for
the engineer but rather permitting a higher fidelity problem to be
solved within a given time constraint. For problems where increasing
fidelity increases the stiffness of the system, there is often a
trade-off between the absolute efficiencies computed for a given
method and the degrading algorithmic efficiencies that arise when
increasing the fidelity. Both time and hardware constraints should be
considered when preparing a parallel calculation.

@@

\subsection{Full-Clip Neumann-Ulam}
\label{subsec:full_clip}

\subsubsection{Weak Scaling}
\label{subsubsec:full_clip_weak}

\subsubsection{Strong Scaling}
\label{subsubsec:full_clip_strong}
